---
title: "Using PurpleAir API"
output: html_notebook
---

# get data from my PA-II sensor on my back deck

```{r}
# packages

#install.packages(c("httr", "jsonlite"))
library(httr)
library(jsonlite)
library(tidyverse)
library(leaflet)
```

# make a request from the PurpleAir API for my sensor

```{r}
# help from https://rdrr.io/cran/httr/f/vignettes/quickstart.Rmd

r <- GET("https://api.purpleair.com/v1/sensors/154517", 
         query = list(api_key = '6B64127F-1D79-11ED-8561-42010A800005'))

r
# rawToChar() will convert raw data 
# to char and store in response variable
response <- rawToChar(r$content)

# parse the json
rawdata = fromJSON(response)

# what was the temp reading? confirm good - seems off from the map value
# and the map widget mentions a conversion - wonder what it is
# found this https://community.purpleair.com/t/the-purpleair-utility/673
# T The temperature inside the sensor’s housing - this temperature reading # will be approximately 8 degrees higher than ambient conditions due to # heat generated by the WiFi module in the sensor’s control board.

# H The humidity inside the sensor’s housing - this humidity reading will # be approximately 4% lower than ambient conditions due to heat generated # by the WiFi module in the sensor’s control board.
rawdata$sensor$temperature - 8
rawdata$sensor$humidity + 4

# correct value to match map - check it
rawdata$sensor$stats$pm2.5_10minute


# to get to the sensor data can build a DF from the sensor data blob
sensordata <- data.frame(rawdata$sensor)
sensordata

# not sure that's useful but?


# example URL for getting a few fields
# https://api.purpleair.com/v1/sensors/37403?sensor_index=37403&fields=pm2.5_30minute%2Ctemperature%2Chumidity

```

# cool, now let's pull some historical data

```{r}
# consult API for more params
# https://api.purpleair.com/#api-sensors-get-sensor-history
# r2 <- GET("https://api.purpleair.com/v1/sensors/154517/history/csv", 
#          query = list(api_key = '6B64127F-1D79-11ED-8561-42010A800005'))
# 
# r2
# fromJSON(rawToChar(r2$content))

# oh, they restricted the access to this
# see https://community.purpleair.com/t/historical-api-endpoints-are-now-restricted/1557

# I can ask for permission but for now I will wait


```

# for PM2.5 need to do a conversion to generate the standard AQI

## note that PurpleAir maps show the 10 min avg value, converted to AQI

```{r}

# going to need a function
# example code in PHP:
# https://community.purpleair.com/t/how-to-calculate-the-us-epa-pm2-5-aqi/877/5

#                                  AQI         RAW PM2.5
	# Good                               0 - 50   |   0.0 – 12.0
	# Moderate                          51 - 100  |  12.1 – 35.4
	# Unhealthy for Sensitive Groups   101 – 150  |  35.5 – 55.4
	# Unhealthy                        151 – 200  |  55.5 – 150.4
	# Very Unhealthy                   201 – 300  |  150.5 – 250.4
	# Hazardous                        301 – 400  |  250.5 – 350.4
	# Hazardous                        401 – 500  |  350.5 – 500.4

# do the calc - what do these variables stand for?
Calc_AQI <- function(Cp, Ih, I1, BPh, BP1) {
  a <- (Ih - I1)
  b <- (BPh - BP1)
  c <- (Cp - BP1)
  return (round((a/b) * c + I1))
}

AQI_from_PM <- function(pm25) {
  # check if valid
  if(pm25 > 0 & pm25 < 1000) {
    if(pm25 > 350.5) return (Calc_AQI(pm25, 500, 401, 500.4, 350.5)) # hazardous
    else if(pm25 > 250.5) return (Calc_AQI(pm25, 400, 301, 350.4, 250.5)) # hazardous
    else if(pm25 > 150.5) return (Calc_AQI(pm25, 300, 201, 250.4, 150.5)) # very unhealthy
    else if(pm25 > 55.5) return (Calc_AQI(pm25, 200, 151, 150.4, 55.5)) # Unhealthy
    else if(pm25 > 35.5) return (Calc_AQI(pm25, 150, 101, 55.4, 35.5)) # Unhealthy for sensitive groups
    else if(pm25 > 12.1) return (Calc_AQI(pm25, 100, 51, 35.4, 12.1)) # Moderate
    else if(pm25 > 0) return (Calc_AQI(pm25, 50, 0, 12, 0)) # good
  }
  else return (NaN) # was not valid pm25
}

# could also structure it as a case_when
# going to use a case_when for the associated health level string for the AQI
healthrating <- function(AQI) {
  return (case_when(is.nan(AQI) ~ "Unknown",
            (0 <= AQI & AQI <= 50) ~ "Good",
            (51 <= AQI & AQI <= 100) ~ "Moderate",
            (101 <= AQI & AQI <= 150) ~ "Unhealthy for Sensitive Groups",
            (151 <= AQI & AQI <= 200) ~ "Unhealthy",
            (201 <= AQI & AQI <= 300) ~ "Very Unhealthy",
            TRUE ~ "Hazardous"
  ))
}



# test it
# Calc_AQI(200,300,201,250.4,150.5)
# AQI_from_PM(200)
# AQI_from_PM(1200)
# working

# add AQI as a column
sensordata <- sensordata %>% mutate(US_EPA_PM25_AQI = AQI_from_PM(stats.pm2.5_10minute), 
                                    HealthLevel = healthrating(US_EPA_PM25_AQI))

# test it
#sensordata$US_EPA_PM25_AQI
#sensordata$HealthLevel
# working!

```

# Should also build a map of my own

### maybe a Leaflet interactive - see the twitter sentiment assignment

### can have a popup with all the stats I care about,

### not just the one value on the purpleair map

# make it interactive to explore tweets from those locations

```{r}

# helper funcs
convert_FtoC <- function(fahrenheit) {
  return (round((5/9)*(fahrenheit-32),2))
}

convert_mbar_to_inHg <- function(mbar) {
  return (round((0.02953)*mbar, 2))
}

# build popup text
# temp and humidity have already been corrected above
popupText = paste0("<b> As of: </b>", as.character(as.POSIXct(sensordata$last_seen, origin = "1970-01-01")), " Pacific<br>",
                    "<b>AQI: </b>", as.character(sensordata$US_EPA_PM25_AQI), "<br>",
                   "<b>AQI Health Rating: </b>", as.character(sensordata$HealthLevel), "<br>",
                   "<b>Temp (&#176;C): </b>", as.character(convert_FtoC(sensordata$temperature)), "<br>",
                    "<b>Temp (&#176;F): </b>", as.character(sensordata$temperature), "<br>",
                   "<b>Pressure (mbar): </b>", as.character(sensordata$pressure), "<br>",
                   "<b>Pressure (inHg): </b>", as.character(convert_mbar_to_inHg(sensordata$pressure)), "<br>",
                    "<b>Humidity (%): </b>", as.character(sensordata$humidity), "<br>",
                   "<b>PM2.5: </b>", as.character(sensordata$stats.pm2.5_10minute), "<br>",
                   "<b>PM1.0: </b>", as.character(sensordata$pm1.0), "<br>",
                   "<b>PM10: </b>", as.character(sensordata$pm10.0), "<br>",
                   "<b>Altitude (ft): </b>", as.character(sensordata$altitude), "<br>"
                    )

site_locations <- leaflet(sensordata) %>%
  addTiles() %>%
  addCircleMarkers(lng = ~longitude, lat = ~latitude, popup = ~popupText,
                   radius = 8, stroke = F, color = "red")

site_locations

```

# and then on to a shiny dashboard
