---
title: "Geospatial - Classify a Raster"
output: html_notebook
---

Working through this course https://www.earthdatascience.org/courses/earth-analytics/ to learn a few things.

# Lesson 5: Classify a Raster

```{r}
library(tidyverse)

getwd()
setwd("C:/Users/tscott/OneDrive - Eastside Preparatory School/courses/DataScience_2/")

options(stringsAsFactors = F)

# libraries for raster data
#install.packages("raster")
#install.packages("rgdal")
library(raster)
library(rgdal)
```

# load dataset
## goal is to classify objects in raster by height

```{r}

lidar_chm <- raster("data/GeoData/earthanalyticswk3/BLDR_LeeHill/outputs/lidar_chm.tif")

# goal is to use 3 classes, what are min and max?
summary(lidar_chm)


# plot histogram and zoom in a bit
hist(lidar_chm,
     main="Distribution of raster cell values in the CHM for LeeHill",
     xlab="Height(m)", ylab="Pixels", col="green",
     xlim=c(2,25),
     ylim=c(0,1000),
     breaks=100)

# plot with specific breaks
hist(lidar_chm,
     main="Distribution of raster cell values in the CHM for LeeHill",
     xlab="Height(m)", ylab="Pixels", col="green",
     breaks=c(0,5,10,15,20,25,30))

```

# create custom breaks based on tree sizes
0-2 = no trees (data below 2 is unreliable in this case)
2-4 = short trees
4-7 = med trees
.> 7 = tall trees

```{r}
hist(lidar_chm,
     main="Distribution of raster cell values in the CHM for LeeHill",
     xlab="Height(m)", ylab="Pixels", col="green",
     breaks=c(0,2,4,7,30))




```

# Map those raster values to new values (reclassify)
0 - 2m = NA
2 - 4m = 1
4 - 7m = 2
. > 7m = 3

```{r}
# create reclass matrix
reclass_df <- c(0,2,NA,
                2,4,1,
                4,7,2,
                7,Inf,3)

reclass_m <-  matrix(reclass_df, ncol=3,byrow=T)

# check it out
reclass_m


# use reclassify()
chm_classified <- reclassify(lidar_chm, reclass_m)

# use barplot to show classes
barplot(chm_classified,
        main="Number of pixels in each height class")

# assign all pixels of class 0 to be NA
chm_classified[chm_classified == 0] <- NA

# they no don't show up
barplot(chm_classified,
        main="Number of pixels in each height class")

# plot the reclassified data in raster view
plot(chm_classified, col= c("red","blue","green"))
# that looks wrong as it is assuming continuous y

# try again
plot(chm_classified,
     legend = F,
     col=c("red","blue","green"),
     axes = F,
     main= "Classified Canopy Height Model \n short, medium, tall trees")

legend("topright",
       legend=c("short trees (2-4m)", "medium trees (4-7m)", "tall trees (>7 m)"),
       fill=c("red","blue","green"),
       border=F,
       bty="n")



```

# be more intentional with colors

```{r}
# create color object with nice new colors!
chm_colors <- c("palegoldenrod", "palegreen2", "palegreen4")

# plot reclassified data
plot(chm_classified,
     legend = FALSE,
     col = chm_colors,
     axes = FALSE,
     # remove the box around the plot
     box = FALSE,
     main = "Classified Canopy Height Model \n short, medium, tall trees")

legend("topright",
       legend = c("short trees", "medium trees", "tall trees"),
       fill = chm_colors,
       border = FALSE,
       bty = "n")


```

# challenge problem - create classified raster showing changes in canopy height (+ and -) before and after the flood

```{r}

# need to have chm from pre-flood and chm from post-flood
# then classify the data as above

# read in the DTM and DSM for pre and post to generate CHM
lidar_pre_dem <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DTM.tif")
lidar_post_dem <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/post-flood/lidar/post_DTM.tif")
lidar_pre_dsm <-  raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DSM.tif")
lidar_post_dsm <-  raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/post-flood/lidar/post_DSM.tif")

# create CHM pre and post
lidar_pre_chm <- lidar_pre_dsm - lidar_pre_dem
lidar_post_chm <- lidar_post_dsm - lidar_post_dem

# now subtract those CHMs to see what difference in canopy height after flood
lidar_resid_chm <- lidar_post_chm - lidar_pre_chm
plot(lidar_resid_chm,
     main = "LIDAR Canopy Changes Post-Flood",
     breaks = c(-20,-10,0,10,20),
     col=c("red","yellow","white","green","darkgreen"))
# maybe a tiny change on the western side?
hist(lidar_resid_chm,
     breaks=50,
     main = "CHM surface elevation value distribution",
     xlab="Elevation (m)", ylab="Frequency", col="green")

# check ranges
summary(lidar_resid_chm)

# lots of NAs - should use barplot
# classify into positive and negative

# create reclass matrix, between -2 and 2m should be NA due to instrument
# sensistivity
reclass_df2 <- c(-Inf,-2,1,
                 -2,2,NA,
                  2,Inf,2)

reclass_m2 <-  matrix(reclass_df2, ncol=3,byrow=T)

# check it out
reclass_m2


# use reclassify()
resid_chm_classified <- reclassify(lidar_resid_chm, reclass_m2)

summary(resid_chm_classified)

# use barplot to show classes
barplot(resid_chm_classified,
        main="Number of pixels in each height class")

# assign all pixels of class 0 to be NA
resid_chm_classified[resid_chm_classified == 0] <- NA

# they no don't show up
barplot(resid_chm_classified,
        main="Number of pixels in each height class")

# two classes, so I guess it is working?
plot(resid_chm_classified)


# need to bucket so it isn't continuous
plot(resid_chm_classified,
     legend = F,
     col=c("red","green"),
     axes = F,
     main= "Classified CHM difference pre vs post-flood")

legend("topright",
       legend=c("negative changes (< -2m)", "positive changes (> 2m)"),
       fill=c("red","green"),
       border=F,
       bty="n")


```
# I guess it worked?

# CROP a raster to clean up edges

```{r}

# need shapefile library
#install.packages("sf")
library(sf)


# open the chm created before
lidar_chm <- raster("data/GeoData/outputs/lidar_chm.tiff")
summary(lidar_chm)

# plot it
plot(lidar_chm,
     col = rev(terrain.colors(50)))

# open vector layer to use for cropping
crop_extent <- readOGR("data/GeoData/earthanalyticswk3/BLDR_LeeHill/clip-extent.shp")


plot(crop_extent, 
     main="Shapefile to use for cropping",
     axes=T,
     border="blue")



# use it to crop and then plot both
lidar_chm_crop <- crop(lidar_chm, crop_extent)
plot(lidar_chm_crop, main="Cropped lidar CHM")
plot(crop_extent, add=T)

# looks nice

```

