---
title: "Geospatial Course Work - LIDAR Data Practice"
output: html_notebook
---

Working through this course https://www.earthdatascience.org/courses/earth-analytics/ to learn a few things.

```{r}
library(tidyverse)

getwd()
setwd("C:/Users/tscott/OneDrive - Eastside Preparatory School/courses/DataScience_2/")

options(stringsAsFactors = F)

# libraries for raster data
#install.packages("raster")
#install.packages("rgdal")
library(raster)
library(rgdal)
```

# Load some data

```{r}
# use raster function to load the tif
# DTM is digital terrain model a.k.a. digital elevation model
# see https://www.earthdatascience.org/courses/earth-analytics/lidar-raster-data-r/lidar-chm-dem-dsm/ for image explanation
lidar_demo <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DTM.tif")

# plot it
plot(lidar_demo, main="Digital Elevation Model - Pre 2013 Flood")




```

```{r}
# zoom in on a section using xlim and ylim
plot(lidar_demo, 
     xlim = c(473000,474000),
     ylim = c(4435000,4435500),
     main="Digital Elevation Model - Pre 2013 Flood")
```

# identify coordinate reference system (CRS) of the data

```{r}
# get CRS properties
crs(lidar_demo)

# save it for later use
myCRS <- crs(lidar_demo)

# what is the extent (what are the boundaries?)
extent(lidar_demo)

# resolution is 1m by 1m since the CRS says +units=m
xres(lidar_demo)
yres(lidar_demo)

```

# inspect the data with plots
```{r}
glimpse(lidar_demo)

# plot a histogram
hist(lidar_demo, breaks=50,
     main = "Distribution of surface elevation values",
     xlab="Elevation (m)", ylab="Frequency", col="green")

# looks pretty bimodal

```

# try a new data file
```{r}

lidar_demo2 <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DSM.tif")

# plot it
plot(lidar_demo2, main="Digital Surface Model - Pre 2013 Flood")

```

```{r}

# get CRS properties
crs(lidar_demo2)

# save it for later use
myCRS <- crs(lidar_demo2)

# what is the extent (what are the boundaries?)
extent(lidar_demo2)

# resolution is 1m by 1m since the CRS says +units=m
xres(lidar_demo)
yres(lidar_demo)
```
# look into the metadata on the geotiff files

```{r}
GDALinfo("data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DSM.tif")

# view extent - @ symbol can extract values from S4 objects
lidar_demo2@extent

# how many layers in the raster object?
nlayers(lidar_demo)


```

# rename the files and start working on a CHM, canopy height model
## CHM = DSM - DTM

```{r}
lidar_dtm <- lidar_demo
lidar_dsm <- lidar_demo2

plot(lidar_dtm, main = "LIDAR Digital Terrain Model")
plot(lidar_dsm, main = "LIDAR Digital Surface Model")
```


```{r}
# now subtract them which gives the residual difference between earth's surface and the ground = heights of trees, etc

lidar_chm <- lidar_dsm - lidar_dtm

# view it
plot(lidar_chm, main = "LIDAR Canopy Height Model")


```

# insert some breaks into the gradient to help things stand out

```{r}

plot(lidar_chm,
     breaks = c(0, 2, 10, 20, 30),
     main = "LIDAR Canopy Height Model",
     col  = c("white","brown","green","darkgreen"))

```

# write this out to save for later

```{r}

# check if output dir exists
#getwd()
if(!dir.exists("data/GeoData/outputs")) {
  dir.create("data/outputs", recursive = T)
}

# export the CHM geotiff file
writeRaster(lidar_chm, "data/GeoData/outputs/lidar_chm.tiff",
            format = "GTiff",
            overwrite = T)


```

# challenge problem - compare pre and post flood data

```{r}

# get all needed files
lidar_pre_dem <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DTM.tif")
lidar_post_dem <- raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/post-flood/lidar/post_DTM.tif")
lidar_pre_dsm <-  raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/pre-flood/lidar/pre_DSM.tif")
lidar_post_dsm <-  raster(x="data/GeoData/earthanalyticswk3/BLDR_LeeHill/post-flood/lidar/post_DSM.tif")


# compare dem (DTM) pre and post flood to compare ground level after flood
lidar_resid_dem <- lidar_post_dem - lidar_pre_dem
plot(lidar_resid_dem,
     main = "LIDAR Elevation Changes Post-Flood")
# maybe a tiny change on the western side?

# create CHM pre and post
lidar_pre_chm <- lidar_pre_dsm - lidar_pre_dem
lidar_post_chm <- lidar_post_dsm - lidar_post_dem

# now subtract those CHMs to see what difference in canopy height after flood
lidar_resid_chm <- lidar_post_chm - lidar_pre_chm
plot(lidar_resid_chm,
     main = "LIDAR Canopy Changes Post-Flood",
     breaks = c(-20,-10,0,10,20),
     col=c("red","yellow","white","green","darkgreen"))
# maybe a tiny change on the western side?
hist(lidar_resid_chm,
     breaks=20,
     main = "CHM surface elevation value distribution",
     xlab="Elevation (m)", ylab="Frequency", col="green")

# a small drop in overall canopy height has the largest peak

```

